The Bolocam Galactic Plane Survey: Survey Description and Data Reduction

### 2009-04-11 methods/ms.tex
Yup.

% Major Sections
% sec:Introduction
% sec:Observations
% sec:FluxCalibration
% sec:Astrometry
% sec:Mapping

@@

\usepackage{rotating}
\usepackage{subfigure}

@@

\def\bgps{BGPS}
\newcommand{\bgpsarea}{150}
\newcommand{\bcamfwhm}{31.2\arcsec}
\newcommand{\ncores}{$10^4$}
\newcommand{\bgpsdepthlow}{20}
\newcommand{\bgpsdepthhigh}{50}

@@

\slugcomment{DRAFT: \today}

@@

\newpage
\pagebreak

\tableofcontents

\newpage
\pagebreak

@@

We present the Bolocam Galactic Plane Survey (BGPS), a 1.1 mm
continuum survey of \bgpsarea\ square degrees of the Galactic Plane in
the 1st, 2nd, and 3rd quadrants.  The BGPS represents the first large
area, systematic survey of the Galactic Plane in the (sub)millimeter
continuum without pre-selected targets.  The survey has detected
\ncores\ clumps to a limiting depth of between \bgpsdepthlow\ and
\bgpsdepthhigh\ mJy RMS.  This paper details the survey observations
and methods.  We treat carefully the comparison of the survey pointing
and flux calibration and and serves as a companion to the public data
release through NASA's Infrared Processing and Analysis Center (IPAC).
We anticipate that the BGPS will be useful as a finder chart for ALMA
and as a long-wavelength complement to {\em Herschel}-HiGAL.

@@

@@

%\section{Overview of the Analysis}
%\label{sec:Overview}

{\it material from the Letter here}

@@

@@

All processed maps are made available through
IPAC\footnote{http://irsa.ipac.caltech.edu/Missions/bolocam.html}.

@@

% first millimeter....
The BGPS is the first (sub)millimeter survey of a substantial fraction
of the Galactic Plane, providing an unbiased look at the high-density
gas most intimately associated with the earliest phases of star
formation.

>>>>

% Major Sections
% sec:Introduction
% sec:Observations
% sec:FluxCalibration
% sec:Astrometry
% sec:Mapping
% sec:Noise
% sec:Photometry
% sec:FinalMaps
% sec:Discussion

@@

\usepackage{rotating}
\usepackage{subfigure}
\usepackage{datetime}
\usepackage{fmtcount}

\newtimeformat{edttime}{\twodigit{\THEHOUR}:\twodigit{\THEMINUTE}~EDT}
\settimeformat{edttime}
\newdateformat{eurodate}{\dayofweekname{\THEDAY}{\THEMONTH}{\THEYEAR} \THEDAY \monthname[\THEMONTH] \THEYEAR}

@@

% -----------------------------------------------------------------------------
% BGPS specific definitions
\def\bgps{BGPS}
\newcommand{\bgpsarea}{150}
\newcommand{\bcamfwhm}{31.2\arcsec}
\newcommand{\ncores}{7628}
\newcommand{\bgpsdepthlow}{20}
\newcommand{\bgpsdepthhigh}{50}
% -----------------------------------------------------------------------------

@@

\slugcomment{DRAFT: \currenttime\ \eurodate\today}

@@

%\newpage
%\pagebreak
%
%\tableofcontents
%
%\newpage
%\pagebreak
%

@@

We present the Bolocam Galactic Plane Survey (BGPS), a 1.1 mm
continuum survey of \bgpsarea\ square degrees of the Galactic Plane in
the 1st, 2nd, and 3rd quadrants.  The BGPS represents the first large
area, systematic survey of the Northern Galactic Plane in the
(sub)millimeter continuum without pre-selected targets.  The survey
has detected \ncores\ clumps to a limiting depth of between
\bgpsdepthlow\ and \bgpsdepthhigh\ mJy RMS.  We include maps of 100
contiguous square degrees from $l=-10.5$ to $l=86.5$ and $b=\pm0.5$
with a resolution of 31\arcsec, as well as regions in IC1396, $l=111$
(NGC7538), W3/4/5, and Gem OB1.  The most striking feature of the maps
is the filamentary structure of cold dust throughout the Galaxy.

This paper details the survey observations and methods.  We treat
carefully the determination of pointing and flux calibration errors
and compare our results to the literature.  Data processing algorithms
that separate astronomical signals from time-variable atmospheric flux
in the data time-stream are discussed.  The algorithms reproduce the
structure of the astronomical sky on a limited range of angular scales
and produce artifacts in the vicinity of bright sources. Specifically,
extended emission smooth on scales larger than about 5\arcmin\ is not
fully recovered in the images.  The spatial frequency response
function of the Bolocam Galactic Plane Survey is discussed.

This paper serves as a companion and guide to the public data release
through NASA's Infrared Processing and Analysis Center (IPAC).  These
images represent an important finder chart for future observations
with ALMA and a long-wavelength spectral band complementary to
imminent submillimeter surveys with SCUBA-2 and Herschel-SPIRE.

@@

Submillimeter observations of Galactic sources have been conducted by
a variety of different groups focusing on small regions since
bolometer technology was developed.  The Submillimeter Common-User
Bolometer Array (SCUBA) on the 15 m James Clerk Maxwell Telescope
mapped many significant regions in the galactic plane over its 8 year
lifetime \citep{holland99,difrancesco08}, but never completed a
contiguous survey of the galactic plane.  Bolocam and Mambo have
previously been used to map large star forming regions at 1.1
\citep{enoch06} and 1.2 mm \citep{motte07}, but again these were
targeted observations.  A recent large area survey with the LABOCA
instrument at 870 \mum\ has recently been reported \citep{schuller09}.

Surveys of the galaxy in CO line emission \citep{dame01,FCRAO} have
revealed much about galactic structure, but do not provide a complete
picture of how star formation follows that structure.  Millimeter
continuum more directly traces the cold dust associated with star
formation.  Continuum data allows for accurate determinations of the
mass of large, dense dust structures when combined with kinematic
distances \citep[e.g.][]{RotationCurve}.

@@

At the distances that we are observing these objects, we are likely
observing aggregates of star forming cores found in a larger molecular
cloud environment, which we will refer to as clumps
\citep{williams00}.

@@

\section{Overview of the Data Reduction}
\label{sec:Overview}

%{\it material from the Letter here}

The data were reduced with a custom pipeline written specifically for
the galactic plane survey based on \citet{enoch06} and
\citet{Laurent2006}.  A key element of the reduction is the iterative
estimation of the atmospheric fluctuations and the astrophysical
signal.  The atmospheric model is developed from a set of principal
components of the bolometer signal time series under the assumption
that the bulk of the correlated signal is atmospheric.  The iterative
estimation process then helps to restore correlated flux due to
astrophysical emission \citep[details in][]{Aguirre2009}.

The Bolocam observations are sensitive to structure between 31\arcsec\
and about 400\arcsec\ (the array field of view).  An indicative
analysis of the spatial filter function of the iterative mapping
scheme is given in \citet{enoch06}.  The data were sampled onto a
uniform 7\farcs2 grid.  Flux density calibrations were performed using
observations of Mars, Uranus, and Neptune, and they are accurate to
$\sim$20\%.

%A calibration scheme capable of
%5\% systematic and $<$5\% statistical errors is being implemented as outlined in
%\citet{Sayers2009}.
%To account for offsets between the observations, each observation was mapped
%individually and aligned to a `master' field that has been tied to
%pointing calibrators. 
Noise properties across the survey are not completely uniform, but the
measured $1\sigma$ level is $\approx43\pm7$ mJy/beam within the
$b=\pm.5$ portion of the plane.
%After the pipeline mapping, the fields were combined using the COMBINE
%command in IRAF with WCS offsets.

The absolute pointing accuracy of the survey has been measured to be
$\sim$6\arcsec\ RMS.  A single $3^\circ \times 1^\circ$ observation of
each region was registered to the pointing calibrators as a `master
field,' and additional observations of that field were aligned to the
master.

Bright and large sources leave an impact on their surroundings when
cleaned with a principal component subtraction.  \object{Sgr B2},
\object{G 34.3+0.15}, and other sources with fluxes $\geq 10$ Jy are
surrounded by substantial negative bowls.  Some flux is restored in
these bowls using the iterative process described in
\citet{Aguirre2009}, but around the brightest sources the impact of
the source on the background estimation is too great to be fully
recovered.

Instrument glitches and cosmic ray hits have been removed by a
combination of outlier rejection in each spatial pixel and hand
flagging of the timestream data.  Streaking in the images results from
incorrect determinations of the zero level across scans when a bright
source is encountered.  Atmospheric noise in observations taken in
poor weather may not be completely removed by the reduction
procedures.
%Particular bolometers may be noisy for a single observation and they may not
%be adequately down-weighted.

@@

All processed maps are made available through
IPAC\footnote{http://irsa.ipac.caltech.edu/Missions/bolocam.html}.
IPAC provides a cutout service for the

@@

% first millimeter....
The BGPS is the first (sub)millimeter survey of a substantial fraction
of the Galactic Plane, providing an unbiased look at the high-density
gas most intimately associated with the earliest phases of star
formation.  Inspection of Figure \ref{fig:BGPSMontage}

Figures 1 and 2 show the results of the BGPS.  Our images show
excellent correspondence with SCUBA maps of small sub-fields
throughout the plane \citep{SCUBAlegacy}.  BGPS contours overlaid on
the GLIMPSE survey \citep{benjamin03} show excellent correlation with
8\mum\ infrared dark clouds (IRDCs).  However, many features in the
BGPS maps do not appear in the Spitzer observations, which suggests
that the 1.1 mm emission traces dense clumps without a background to
absorb.  A study of IRDC correlation with BGPS emission is underway
\citep{Battersby2009}.

% implies star forming....

Filamentary structures with aspect ratios $\geq 10$ are present
throughout the galactic plane.  The most crowded fields consist of
frothy and clumpy structure.  There are 5000-10000 distinct sources,
the majority of which are at least moderately resolved
\citep{rosolowsky09}.  We suspect that higher resolution observations
will resolve BGPS clumps into clusters of protostellar objects, as has
been seen in Serpens \citep{enoch08,Testi1998} and OMC 1
\citep{Beuther2004,johnstone99} where bolometer observations at low
resolution have been complemented by interferometer observations at
high resolution.

There are clear maxima in the emission filling factor in the Galactic
Center and in the region from $l=23$ to $l=31$, the line of sight
through the densest part of the molecular ring.  Additional areas of
high source density are seen at $l=10$ and $l=13$.  The filling factor
of 1.1 mm continuum emission at the 100 mJy level is $\sim 20$\% per
square degree from $l=-10$ to $l=20$ and $\sim 10$\% from $l=20$ to
$l=50$.
%In the range
%$l=50$ to $l=75$, the filling factor is below 1\% and very few sources have
%been detected.  
This result is in contrast to CO filling factors $\sim1$ in the inner galactic
plane \citep{dame01,FCRAO}, confirming that 1.1 mm continuum emission traces
denser material than CO.  
%The small filling factor of the 1.1 mm emission
%compared to CO emission  shows that the continuum is directly associated with
%star forming material, whereas the CO also traces less dense material not
%directly associated with star formation.

In some cases, distance to BGPS clumps can be determined by matching
Galactic Ring Survey $^{13}$CO and 1.1 mm continuum morphology
\citep{IRDCdistance,FCRAO}.  However, in other cases the association
between the CO and 1.1 mm data is not clear, either because of
confusion or because the BGPS sources are too compact to identify in
$^{13}$CO morphology.  Heterodyne follow-up observations using dense
gas tracers (NH$_3$, N$_2$H$^+$, HCO$^+$, and CS) are being conducted
to provide radial velocity, density, chemistry, and temperature
measurements \citep{schlingman09, battersby09, nordhaus09}.

BGPS clumps are associated with the densest portions of giant
molecular clouds (GMCs) traced by CO surveys
\citep[e.g.,][]{dame01,FCRAO}.  The brightest clumps visible in our
maps, e.g. \object{Sgr B2}, \object{G34.3+0.15}, \object{W 51},
\object{W 43}, \object{W 49}, and \object{M 17}, correspond to well
known star forming regions.
%(the saturation is only a display artifact; the astrophysical sources are
%always much fainter than the atmosphere and therefore do not saturate the
%detectors).  
These associations suggest that other bright sources of millimeter
emission will prove to be massive star or cluster forming regions.
